---
title: "The Crop Analytics Orchestra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Crop Analytics Orchestra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS arm64 (darwin 25.5.0),
     R 4.5.2, one core. No install and no external gate script runs live in
     this vignette, so the time is dominated by knitr overhead and the
     load-check of the fifteen roster members already installed here. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  dpi = 150
)
```

```{r library}
library(cropOrchestra)
```

## Why

I have R installed, a nitrogen-rate decision to make, and a colleague's word
that an "orchestra" of packages can carry it from raw paddock data to a
defensible number; there are fifteen of them, several are not on CRAN, and
before I trust any of it I want to know what is actually here, what each
piece is for, and whether I can check that for myself rather than take it on
faith. What is `cropOrchestra`, and how do I find out what I have? This is
the question an agronomist, analyst or breeder asks the first time somebody
hands them the Crop Analytics Orchestra rather than a single package: not
"how does the mathematics work" but "what is actually here, and can I trust
the list". `cropOrchestra` answers it directly. It is not itself an
analytical package -- it ships no models -- it is the front door: a roster,
a version checker, an installer, and a gate runner over the released
members. The other packages in the orchestra take a grain-production
question -- which nitrogen rate, which variety, when to test, when to
harvest -- from raw data to a recorded decision, and refuse rather than
guess where the evidence does not support one; every hand-off between them
is a typed manifest, every abstention is a typed refusal rather than a
silent `NA`, and every external fact a package relies on carries provenance
back to its source. When the whole chain was run end to end on a synthetic
barley belt (Study A0), it returned a headline number that beat a naive
baseline by a wide margin; this vignette does not rerun that chain, so the
number itself, and the sentence that says it is not recomputed here, are
held for *Limits*. What this vignette does show, on this machine, right
now, is what `cropOrchestra` itself can check without running a single crop
model: which packages are on the release roster, which of them are actually
installed, and whether the two agree.

## What

`cao_roster()` returns the fifteen-member release roster frozen on
2026-09-02: nine *core* members that every study routes through, and six
*supporting* members that particular studies need, each row carrying its
tier, its necessity-register role, the `max578` repository it publishes
through, and whether that repository is public yet. A member earns its row
by answering four questions in the release's necessity register: which crop
question needs it, what the nearest tool outside the orchestra is, why that
outside tool does not suffice -- one of four admissible reasons: a missing
capability, a contract the pipeline requires that the outside tool cannot
emit, an independent-oracle grounding it does not carry, or a translation
layer that does not exist outside -- and where it was load-bearing in a
study, an arbitration, or a fleet gate. The method statement behind every
member is the same three-step discipline: the orchestra *picks* the best
available method, old or new, its own or external, *adopts or adapts* it
behind the typed-manifest contract, and *translates* it into a grower- or
breeder-facing decision. A member that only adapts an external tool --
wrapping it behind the contract rather than adding new mathematics -- is
justified by the contract alone, and the register says so plainly rather
than overstating its novelty.

`cao_versions()` joins that fixed roster with what this R session can
actually load: an `installed_version` column read through
`requireNamespace()` rather than a hard `library()` call, so a member whose
repository is still private is reported as not installed rather than
raising an error, plus a `match` column recording whether the installed
version equals the frozen one.

`cao_status()` turns that join into a report: every member classified as
installed and matching, installed but drifted, or missing, and for every
missing member, a stated reason -- its repository is not yet public, or it
is public but not yet mirrored on `max578.r-universe.dev`.

`cao_install()` installs a tier (`"core"`, `"supporting"`, `"all"`) or an
explicit set of package names, one member at a time, from the `max578`
r-universe mirror and CRAN, so one member's still-private repository never
stops the rest from installing; `dry_run = TRUE` resolves and returns the
plan without installing anything, which is what the next section runs.

`cao_check()` goes one step further than a status report: pointed at a
local ORCHESTRA leader workspace, it runs the fleet's own gate scripts --
the cross-member refusal contract and the manifest conformance suite -- and
reports pass or fail for each; without a workspace, it falls back to
`cao_status()` plus a load check of every member that is installed.

## Do

### The roster, as frozen

```{r roster}
roster <- cao_roster()
nrow(roster)
attr(roster, "freeze_date")
```

```{r roster-kable, echo = FALSE}
knitr::kable(
  roster[, c("member", "tier", "role")],
  col.names = c("Member", "Tier", "Role"),
  caption = paste0(
    "The frozen roster (", nrow(roster), " members, freeze ",
    attr(roster, "freeze_date"), ")."
  )
)
```

### What is actually installed on this machine

`cao_status()` prints its own report, so the table below is read straight
from the console, not retyped.

```{r status}
status <- cao_status()
```

```{r fig-roster-status, echo = requireNamespace("ggplot2", quietly = TRUE), eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Every roster member at the 2026-09-02 freeze, grouped by tier and coloured by whether the version installed on this machine matches the frozen release, has drifted ahead of it, or is not installed here; the label in each cell reads frozen version / installed version."}
plot_df <- status$versions
plot_df$member <- factor(plot_df$member, levels = rev(plot_df$member))
plot_df$tier_label <- ifelse(
  plot_df$tier == "core",
  paste0("core (", sum(plot_df$tier == "core"), ")"),
  paste0("supporting (", sum(plot_df$tier == "supporting"), ")")
)
plot_df$status_label <- gsub("_", " ", plot_df$status, fixed = TRUE)
ggplot2::ggplot(
  plot_df, ggplot2::aes(x = 1L, y = member, fill = status_label)
) +
  ggplot2::geom_tile(width = 0.9, height = 0.85, colour = "white") +
  ggplot2::geom_text(
    ggplot2::aes(label = paste0(
      version, " / ",
      ifelse(is.na(installed_version), "-", installed_version)
    )),
    colour = "white", size = 3
  ) +
  ggplot2::facet_grid(tier_label ~ ., scales = "free_y", space = "free_y") +
  ggplot2::scale_fill_manual(
    name = "status",
    values = c(
      "installed matching" = "#0072B2",
      "installed drifted" = "#D55E00",
      "missing" = "#999999"
    )
  ) +
  ggplot2::labs(
    x = NULL, y = NULL,
    title = "Roster status at the 2026-09-02 freeze, this machine"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    strip.text.y = ggplot2::element_text(angle = 0)
  )
```

```{r fig-roster-status-skip, echo = FALSE, eval = !requireNamespace("ggplot2", quietly = TRUE), results = "asis"}
cat("ggplot2 is not installed; the roster status figure is skipped.\n")
```

### Planning an install without touching anything

```{r install-plan}
plan <- cao_install(members = "core", dry_run = TRUE)
```

```{r install-plan-kable, echo = FALSE}
knitr::kable(
  plan,
  col.names = c("Member", "Repositories tried"),
  caption = paste0(
    "The resolved install plan for the core tier (", nrow(plan),
    " members); dry run, nothing installed."
  )
)
```

```{r install-real, eval = FALSE}
# Installs whatever in the plan above is not already present. Skipped here
# because rendering this vignette must never modify the R library it runs
# in; run this line yourself against your own library to install the core
# tier for real.
cao_install(members = "core")
```

### Checking the fleet's own gates

```{r check-fallback}
check_fallback <- cao_check()
check_fallback$n_loadable
check_fallback$n_load_failed
```

```{r check-workspace, eval = FALSE}
# Needs a local ORCHESTRA leader workspace (not shipped with this package)
# whose integration/ directory holds the fleet's gate scripts. Skipped here
# because the path below is specific to the machine this vignette was
# written on, not portable to a reader's own layout.
cao_check(workspace = "/path/to/ORCHESTRA_dev")
```

## Read

```{r read-prep, echo = FALSE}
n_core <- sum(roster$tier == "core")
n_supporting <- sum(roster$tier == "supporting")
n_private <- sum(roster$visibility == "private")
pct_private <- round(100 * n_private / nrow(roster))
drifted <- status$versions[status$versions$status == "installed_drifted", ]
drifted_desc <- paste0(
  drifted$member, " (frozen ", drifted$version, ", installed ",
  drifted$installed_version, ")"
)
n_installed <- nrow(roster) - status$n_missing
```

```{r read-text, echo = FALSE, results = "asis"}
cat(
  "Of the ", nrow(roster), " roster members, ", n_core, " are core and ",
  n_supporting, " are supporting; ", n_private, " of the ", nrow(roster),
  " (", pct_private, " per cent, roughly half) are on a private ",
  "repository at the freeze date, and `cao_status()` says so plainly, ",
  "member by member, rather than presenting a partial install as if the ",
  "roster installs in full today.\n\n",
  "On the machine this vignette was rendered on, ", status$n_matching,
  " members match their frozen version exactly and ", status$n_missing,
  " are missing. ",
  if (status$n_drifted > 0L) {
    paste0(
      status$n_drifted,
      if (status$n_drifted == 1L) " member has" else " members have",
      " drifted ahead of its frozen version rather than behind it: ",
      paste(drifted_desc, collapse = ", "), ".\n\n"
    )
  } else {
    "None has drifted from its frozen version.\n\n"
  },
  "Of the ", n_installed, " installed members here, ",
  check_fallback$n_loadable, " load cleanly through ",
  "`requireNamespace()` and ", check_fallback$n_load_failed, " do not. ",
  "The figure shows the same three-way split by tier: a reader can see at ",
  "a glance whether drift, where it exists, clusters in the core tier ",
  "every study routes through or in the supporting tier particular ",
  "studies need, which the roster table above does not show without ",
  "reading all ", nrow(roster), " rows.\n",
  sep = ""
)
```

## Limits

This vignette shows what `cropOrchestra` itself can determine without
touching a crop model: a roster, a version comparison, and a load check. It
does not run `apsimR`, does not fit anything through `kernR` or
`flexyBayes`, and does not price a decision through `decideR` -- it cannot
speak to whether any member's science is correct, only to whether the
member is present and at the frozen version. The install call above is a
dry run for the same reason `cao_check(workspace = ...)` above is not
evaluated: rendering this vignette must never modify the R library it runs
in, and the workspace path is specific to the machine it was written on
rather than portable to a reader's own layout; a reader who wants either
result runs the commands shown, verbatim, against their own library and
their own copy of the ORCHESTRA leader workspace.

The headline claim behind the necessity register -- that when the roster's
core chain (`apsimR` + `quorum` + `PESTO` as the mechanism lens, `kernR`'s
TACI as the field-agreement lens, `decideR` as the decision lens, `optimix`
as the site-value lens, and `grainPlan` as the grower-facing translation)
was run end to end on a synthetic barley belt, the orchestra's cascade lost
2,960 AUD per hectare against the planted truth over 198 decision cells,
against 5,818 AUD per hectare for a naive mechanism-only arm -- 49 per cent
less loss, reproducing on a re-seeded belt -- is not recomputed in this
vignette; it is cited to Study A0's own ledger (`studies/A0/REPORT.md` and
`SUMMARY.md` in the ORCHESTRA leader workspace) and should be read there,
not trusted from this paragraph alone. It is a controlled recovery exercise
on a synthetic belt, not a real one: the shape of the result has real-data
cover -- the belt's nitrogen-response slopes partially bracket a real
barley multi-environment trial -- but the magnitudes do not, and they
should never be quoted as a field result. `cao_status()` and `cao_check()`
exist precisely so a claim like this one is always read alongside what is
actually installed and actually gated on the machine making the claim, not
asserted from memory.

Everything numeric in *Read* above describes one machine at one moment: a
different library, a package upgraded after this vignette was rendered, or
the roster's next freeze would all change the counts without changing
anything this vignette asserts about how to find them out.

## What to read next

*A quorum of lenses on Study A0* (in `quorum`) resolves the same 200
recorded decision cells the headline number above comes from, lens by
lens, and is the closest a reader can get to the cascade's mechanics
without rerunning the study. *From an inference result to a grain plan*
(in `grainPlan`) picks up exactly where this front door leaves off, turning
a manifest-typed inference result into a grower-facing plan. *Getting
started with decideR* introduces the decision lens the headline cascade
above relies on.

## Reproduce

This vignette draws no random numbers and calls no external service: every
table and figure is a deterministic function of the roster shipped inside
this package and of what is installed in the R library it is rendered
against, which is why the counts above are read live rather than typed by
hand.

```{r session-info}
sessionInfo()
```
