---
title: "The Crop Analytics Orchestra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Crop Analytics Orchestra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What this is

The Crop Analytics Orchestra is a set of purpose-built R packages, published
under the `max578` namespace, that together take a grain-production question
-- which nitrogen rate, which variety, when to test, when to harvest -- from
raw data to a recorded decision, and refuse rather than guess where the
evidence does not support one. Every hand-off between packages is a typed
manifest, every abstention is a typed refusal rather than a silent `NA`, and
every external fact a package relies on carries provenance back to its
source. `cropOrchestra` is not itself an analytical package -- it ships no
models -- it is the front door: a roster, a version checker, an installer,
and a gate runner over the released members.

## The released roster

Fifteen packages are released at the 2026-09-02 freeze: nine *core* members
that every study routes through, and six *supporting* members that
particular studies need. `cao_roster()` returns this table as a data.frame.
It is reproduced here for reading.

```{r}
cropOrchestra::cao_roster()
```

| Member | Tier | Role |
|---|---|---|
| flexyBayes | core | the inference hub |
| PESTO | core | simulator inversion + ensemble UQ |
| kernR | core | the TACI engine (flagship) |
| apsimR | core | the mechanism member |
| flexyBayesOrchestra | core | the composition layer |
| terroir | core | the grounded covariate source |
| masque | core | the sovereignty bridge |
| decideR | core | the decision closer |
| grainPlan | core | the practitioner last mile |
| gpfield | supporting | change-of-support spatial GP |
| kalmix | supporting | state-space + ACI |
| koine | supporting | the fourth opinion |
| proxymix | supporting | the mixture calculus |
| optimix | supporting | the optimisation meta-layer |
| cdzoo | supporting | the causal-discovery harness |

Roughly half the roster is on a private repository at the freeze date --
`cao_status()` says so honestly, member by member, rather than pretending the
roster installs in full today.

## The necessity principle

Every member in the table above earned its place by answering four
questions in the release's necessity register: (1) which crop question
needs it, (2) what the nearest tool outside the orchestra is, (3) why that
outside tool does not suffice -- one of four admissible reasons: a missing
capability, a contract the pipeline requires that the outside tool cannot
emit, an independent-oracle grounding it does not carry, or a translation
layer that does not exist outside -- and (4) where it was load-bearing in a
study, an arbitration, or a fleet gate. The method statement behind every
member is the same three-step discipline: the orchestra *picks* the best
available method, old or new, its own or external, *adopts or adapts* it
behind the typed-manifest contract, and *translates* it into a grower- or
breeder-facing decision. A member that only adapts an external tool --
wrapping it behind the contract rather than adding new mathematics -- is
justified by the contract alone, and the register says so plainly rather
than overstating its novelty.

## The headline result, read honestly

Study A0 chains the orchestra end to end on a synthetic barley belt: forty
sites on real Australian grain-belt coordinates with real climate and soil
covariates, a hidden subsoil constraint at twelve of them, and a mechanism
lens (`apsimR` + `flexyBayesOrchestra` + `PESTO`), a field-agreement lens
(`kernR`'s TACI), a decision lens (`decideR`), a site-value lens (`optimix`),
and a grower-facing translation (`grainPlan`). The headline number: the
orchestra's cascade loses 2,960 AUD/ha against the planted truth over 198
decision cells, versus 5,818 AUD/ha for a naive mechanism-only arm -- 49 per
cent less loss, and the result reproduces on a re-seeded belt.

This is a controlled recovery exercise on a synthetic belt, not a real one.
The shape of the result has real-data cover -- the belt's
nitrogen-response slopes partially bracket a real barley multi-environment
trial -- but the magnitudes do not, and they should never be quoted as a
field result. `cao_status()` and `cao_check()` exist precisely so a claim
like this one is always read alongside what is actually installed and
actually gated, not asserted from memory.

## Install, status, check

Installing the roster never fails as a whole because one member's repository
happens to be private today. Each member's outcome is reported honestly.

```{r}
# Plan only -- no packages are touched.
cropOrchestra::cao_install(members = "core", dry_run = TRUE)

# Install the core roster from the max578 r-universe mirror, falling back
# to CRAN; report success, failure, or absence per member.
cropOrchestra::cao_install(members = "core")
```

`cao_versions()` joins the roster with what is actually loadable in this R
session, and `cao_status()` turns that into a report that names, for every
missing member, whether its repository is not yet public or simply not yet
mirrored on r-universe.

```{r}
cropOrchestra::cao_versions()
cropOrchestra::cao_status()
```

`cao_check()` goes one step further: pointed at the ORCHESTRA leader
workspace, it runs the fleet's own gate scripts (the refusal contract and
the manifest conformance suite) and reports pass or fail for each. Without a
workspace, it falls back to `cao_status()` plus a load-check of every member
that is installed.

```{r}
cropOrchestra::cao_check(
  workspace = "/path/to/ORCHESTRA_dev"
)

# No workspace: status plus a load-check of whatever is installed.
cropOrchestra::cao_check()
```
